---
title: "Expression patterns"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
theme_set(theme_bw(base_size = 14))
require(tidyverse)
require(doMC)
registerDoMC(detectCores()-1)
```

## Load expression data
```{r}
DATADIR="~/scailscratch/dox/"
library(data.table)
source("../code/utils.R")

source("../code/load_data.R")

input=quantile_normalize(input)
```

## Run basic DE analysis

```{r}
anno$dbgap=as.character(dbgap[anno$individual])
genes=rownames(input)
pvalues=foreach(gene=genes, .errorhandling='pass', .combine = bind_rows) %dopar% {
  y=input[gene,]
  data.frame(gene=gene,
             lm_p=anova( lm(y ~ as.factor(anno$conc)), lm(y ~ 1))["Pr(>F)"][2,1],
             spearman_p=cor.test(y, anno$conc, method="spearman")$p.value,
             stringsAsFactors = F
  )
}
qplot(pvalues$lm_p, geom="blank") + geom_histogram() + theme_bw(base_size = 16) + xlab("p-value")
#ggsave("../figures/de_boring.pdf", width = 5, heigh=5)
q=p.adjust(pvalues$lm_p,method="BH")
cat( sum(q<.05), "/", length(q), "(", mean(q<.05), ")\n") # 98% DE! 

q=p.adjust(pvalues$spearman_p,method="BH")
cat( sum(q<.05), "/", length(q), "(", mean(q<.05), ")\n") # 98% DE! 
```

## Expression pattern analysis
```{r}

cached_fit_fn="../data/expression_patterns.RData"
if (F) {
  require(rstan)

  dim(input)
  y=t(scale(t(input)))
  x=model.matrix(~as.factor(conc)-1, data=anno)
  #x=x[,2:ncol(x)]
  data=list(y=y, N=ncol(y), D=nrow(y), x=x, P=ncol(x), K=20)

  mixlm_model=stan_model("mixlm_pred.stan")
  mixlm_fit=optimizing(mixlm_model, data, verbose=T, iter=100,  as_vector=F) # takes a while... 
  save(file=cached_fit_fn, mixlm_fit)
} else {
  load(cached_fit_fn)
}
ggplot( data.frame(n=1:20, p=sort(mixlm_fit$par$p, decreasing = T)), aes(n,p)) + geom_bar(stat="identity") + theme_bw(base_size = 16) + ylab("Mixture weight") + xlab("Mixture component")
```

Take top six clusters
```{r}
to_use=order(mixlm_fit$par$p, decreasing = T)[1:6]
mix_means=mixlm_fit$par$betas[to_use,]

assignments=apply(mixlm_fit$par$logprob, 1, which.max)

plot(table(assignments))
```

Plot expression patterns
```{r}
concs=sort(unique(anno$conc))
colnames(mix_means)=concs

require(reshape2)
m=melt(mix_means, value.name="mean")
colnames(m)=c("mixture","conc","mean")
weights=mixlm_fit$par$p[to_use]
m$pi=weights[m$mixture]
m$mixture=as.factor(m$mixture)
mixture_centroids = ggplot(m, aes(conc, mean, col=mixture, shape=mixture, size=pi*100)) + geom_line(alpha=.7) + geom_point() + scale_x_sqrt(breaks=concs, labels=concs, expand=c(0,0)) + theme_bw(base_size = 14) + ylab("Relative expression") + xlab("Doxorubicin concentration") + scale_size("% genes") + scale_color_manual("Cluster ID", values=cbPalette) + scale_shape("Cluster ID")
mixture_centroids
#ggsave("../figures/mixture_centroids.svg",width=5.5,height=4)
#ggsave("../figures/mixture_centroids.pdf",width=5.5,height=4,device=cairo_pdf)
```

1. Down regulated
2. Initial increase, then decrease
3. Up regulated
4. Transient down regulation
5. Transient up regulation
6. Down regulation and then some recovery. 


```{r}
source("GOtest.R")

to_use=order(mixlm_fit$par$p, decreasing = T)[1:6]
mix_means=mixlm_fit$par$betas[to_use,]

assignments=apply(mixlm_fit$par$logprob[,to_use], 1, which.max)
save(file="../data/expression_patterns.RData", mixlm_fit, assignments)


GO_res=foreach(i=1:6) %do% {
  summary(GOtest(rownames(y), rownames(y)[assignments==i], ontology = "BP")$GO$BP, pvalue = 1e-5)
}

go_data <- goHeatmapData(GO_res)

go_heatmap <- plotHeatmap(go_data, labCol = "")

```

```{r}
gene_cluster_dir="../data/gene_clusters/"
dir.create(gene_cluster_dir)
foreach(i=1:6) %do% {
  write.table(rownames(input)[assignments==i], paste0(gene_cluster_dir,i,".txt"), col.names = F, row.names = F, quote=F)
}
```

Ran HOMER on dag-ws-4

```{r}

all_go=foreach(i=1:6, .combine = bind_rows) %do% {
homer_res_dir=paste0(gene_cluster_dir,"homer",i,"/")
  all_txt=list.files(homer_res_dir, glob2rx("*.txt"))
  to_load=setdiff( all_txt, c("knownResults.txt","motifFindingParameters.txt","targetIDs.ug.txt"))
  foreach(fn=to_load, .combine = bind_rows) %do% {
    fread(paste0("cut -f1-9 ",homer_res_dir,fn), data.table = F) %>% mutate(source=fn, TermID=as.character(TermID))
  } %>% mutate(cluster=i)
}


all_go = all_go %>% filter(`Genes in Term` < 2000 )
```

```{r}
all_go_sub = all_go %>% filter(`Genes in Term` > 50 )
interesting_terms = all_go_sub %>% group_by(cluster) %>% top_n(5, -logP) %>% .$TermID %>% unique()

clusters=c("1.Down","2.Up-DOWN","3.Up","4.Down-up","5.Up-down","6.DOWN-up")


#all_go = all_go %>% mutate(q=p.adjust(Enrichment, method="BH"))

all_go_sub = all_go_sub %>% mutate(OR=(`Target Genes in Term` / `Genes in Term`) / ((`Total Target Genes` - `Target Genes in Term`) / (`Total Genes` - `Genes in Term`) ) )


go_mat = all_go_sub %>% filter(TermID %in% interesting_terms) %>% select(Term, cluster, OR) %>% spread(cluster, OR) 
rownames(go_mat) = go_mat$Term
go_mat$Term=NULL
go_mat = as.matrix(go_mat)
heatmap(go_mat)
```

```{r}
row_order_i = go_mat %>% dist() %>% hclust() %>% as.dendrogram() %>% order.dendrogram()
row_order=rownames(go_mat)[row_order_i]
col_order = go_mat %>% t() %>% dist() %>% hclust() %>% as.dendrogram() %>% order.dendrogram()

all_go_sub %>% filter(TermID %in% interesting_terms) %>% 
  mutate(label=paste0(format(OR, digits=1),ifelse(q<0.1,"*","")), 
         cluster=factor(cluster, col_order, clusters[col_order]), 
         Term=factor(Term, row_order)) %>% 
   ggplot(aes(cluster, Term, fill=OR, label=label)) + geom_tile() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + scale_fill_gradient(low="white",high="black") + geom_text() + xlab("Cluster") + ylab("Gene set") # + scale_fill_gradient2(low = "white", mid = "yellow", high="red", midpoint = 2)
ggsave("../figures/cluster_go.pdf",height=6,width=9,device=cairo_pdf)
```

## Motifs

```{r}
all_tf=foreach(i=1:6, .combine = bind_rows) %do% {
  homer_res_dir=paste0(gene_cluster_dir,"homer",i,"/")
  temp=fread(paste0(homer_res_dir,"knownResults.txt"), data.table = F) 
  colnames(temp)[6:9]=c("ntarget_w_motif","ptarget_w_motif","nback_w_motif","pback_w_motif")
   temp %>%  mutate(cluster=i)
} %>% mutate(ptarget_w_motif=gsub("%","",ptarget_w_motif) %>% as.numeric(), 
                   pback_w_motif=gsub("%","",pback_w_motif) %>% as.numeric(), 
                  OR=ptarget_w_motif / pback_w_motif,
                  motif=str_split_fixed(`Motif Name`,"/",3)[,1]) 
```

```{r}
#%>% filter(`q-value (Benjamini)` < 0.1)
top_tf = all_tf %>% filter(`q-value (Benjamini)` < 0.4)  %>% group_by(cluster) %>% top_n(10, -`P-value`) %>% ungroup()
dim(top_tf)
interesting_tf=unique(top_tf$motif)
plot_tf = all_tf %>% 
  filter(motif %in% interesting_tf) %>%
  group_by(cluster,motif) %>%
  summarize(OR=mean(OR),q=min(`q-value (Benjamini)`),Consensus=Consensus[which.min(`q-value (Benjamini)`)]) %>%
  ungroup() %>%
  mutate(motif=str_split_fixed(motif,"[()]",2)[,1])

tf_mat = plot_tf %>% 
  select(cluster, OR, motif) %>% 
  spread(cluster, OR) %>%
  as.data.frame()
rownames(tf_mat)=tf_mat$motif
tf_mat$motif=NULL
tf_mat=as.matrix(tf_mat)
motif_hc = tf_mat %>% dist() %>% hclust() 
plot(motif_hc)
motif_order = motif_hc %>% as.dendrogram() %>% order.dendrogram()
motif_names_order=rownames(tf_mat)[motif_order]

consensus_ordered = plot_tf %>% group_by(motif) %>% summarise(Consensus=Consensus[which.min(q)]) %>% right_join(data.frame(motif=motif_names_order, stringsAsFactors=F),by="motif") %>% .$Consensus

plot_tf %>%
  mutate(cluster=factor(cluster,1:6,clusters),
         motif=factor(motif,motif_names_order) %>% as.numeric) %>%
  ggplot(aes(cluster,motif,fill=log(OR),label=ifelse(q<0.05,"**",ifelse(q<0.5,"*","")))) + geom_tile() + scale_fill_gradient2() + geom_text() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), legend.position = "right") + scale_y_continuous(breaks=seq_along(motif_names_order),labels=motif_names_order,sec.axis=dup_axis(name="Consensus motif",labels=consensus_ordered)) + xlab("Cluster") + ylab("Transcription factor") + coord_cartesian(ylim=c(1.5,length(motif_names_order)-0.5))
ggsave("../figures/cluster_tf.pdf",height=6,width=8,device=cairo_pdf)
```

- ZNF143 transcription factor mediates cell survival through upregulation of the GPX1 activity in the mitochondrial respiratory dysfunction https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3542592/
- YY1 has some involvement in DNA repair and apoptosis. http://www.genecards.org/cgi-bin/carddisp.pl?gene=YY1 Expression and regulation of Cxcl12 during oxidative stress in pancreatic beta cells. http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0059679
- Ronin: regulation of mitochondria and oxidative stress. https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4767647/
- Mef2d myocyte enhancer factor 2. Stress response and apoptosis. http://www.genecards.org/cgi-bin/carddisp.pl?gene=MEF2D
