---
title: "Sequencing decisions"
date: 2016-06-14
output:
  html_document:
    toc: true
    toc_float: true
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(fig.width = 8, fig.heigth = 8, cache = FALSE)
```

We have sequenced each library once.
Now we need to determine:

* Which samples require more sequencing?
* Which master mixes need rebalanced?

## Setup

```{r packages, message=FALSE}
library("DT")
library("dplyr")
library("ggplot2")
library("cowplot")
```

Input annotation.

```{r anno}
anno <- read.delim("../data/annotation.txt", stringsAsFactors = FALSE)
dim(anno)
```

Input mapping summary statistics.
229 of the 230 samples have been mapped.

```{r mapping}
mapping <- read.delim("../data/star_output.tsv", check.names = FALSE,
                      stringsAsFactors = FALSE)
dim(mapping)
```

Split filename into id, flow cell, and lane.

```{r split-filename}
fname_parts <- strsplit(mapping$Sample, "-")
fname_df <- t(as.data.frame(fname_parts))
mapping <- mapping %>%
  mutate(id = paste(fname_df[, 1], fname_df[, 2], fname_df[, 3], sep = "-"),
         flow_cell = fname_df[, 4],
         lane = fname_df[, 5])
```

Merge the two data sets.

```{r merge}
d <- merge(anno, mapping, by = "id")
dim(d)
datatable(d)
```

Rename the columns.
The spaces were troublesome, even when using `ggplot::aes_string`.

```{r rename-cols}
colnames(d)[16:21] <- c("fname", "speed", "raw", "length", "unique",
                        "unique_perc")
colnames(d)
```

Scale the counts to millions.

```{r millions}
d <- d %>%
  mutate(raw_mil = raw / 10^6,
         unique_mil = unique / 10^6)
```

## How many reads per lane on the HiSeq 4000?

```{r summarize-per-lane}
d_lane <- d %>%
  group_by(flow_cell, lane) %>%
  summarize(raw_lane = sum(raw_mil),
            unique_lane = sum(unique_mil))
```

The HiSeq 4000 generated a mean of `r round(mean(d_lane$raw_lane), digits = 0)` +/- `r round(sd(d_lane$raw_lane) / length(d_lane$raw_lane), digits = 1)` (SEM) million raw reads per lane, which converted to a mean of `r round(mean(d_lane$unique_lane), digits = 0)` +/- `r round(sd(d_lane$unique_lane) / length(d_lane$unique_lane), digits = 1)` (SEM) million uniquely mapping reads.

```{r plot-per-lane}
p_lane <- ggplot(d_lane, aes(x = raw_lane, y = unique_lane, color = flow_cell)) +
  geom_point() +
  scale_color_discrete(name = "Flow cell") +
  labs(x = "Number of raw reads (millions)",
       y = "Number of uniquely mapping reads (millions)",
       title = "HiSeq 4000 output per lane")
p_lane
```

The variation in the two metrics does not appear to segregate strongly by flow cell.

## What is the variation in sequencing across samples?

The differences in the number of raw sequences and the conversion to mapped reads is fairly consistent across the master mixes.

```{r plot-per-sample}
p_sample <- p_lane %+% d %+% aes(x = raw_mil, y = unique_mil) +
  labs(title = "Sequencing per sample") +
  facet_wrap(~master_mix)
p_sample
```

Only `r sum(d$raw_mil > 45)` samples have greater than 45 million raw reads
(range: `r round(min(d$raw_mil), digits = 2)` - `r round(max(d$raw_mil), digits = 2)` million).
Only `r sum(d$unique_mil > 30)` samples have greater than 30 million uniquely mapping reads
(range: `r round(min(d$unique_mil), digits = 2)` - `r round(max(d$unique_mil), digits = 2)` million).

```{r range-per-sample}
p_raw <- ggplot(d, aes(x = reorder(id, raw_mil), y = raw_mil)) +
  geom_point() +
  geom_hline(aes(yintercept = 45), color = "red") +
  theme(axis.text.x = element_blank()) +
  labs(x = "Sample",
       y = "Raw reads (millions)",
       title = "Raw reads")
p_unique <- p_raw %+% aes(y = unique_mil) +
  geom_hline(aes(yintercept = 30), color = "blue") +
  labs(x = "Sample",
       y = "Number of uniquely mapping reads (millions)",
       title = "Uniquely mapping reads")
plot_grid(p_raw, p_unique)
```

## What is relationship between library concentration estimates and sequencing?

The goal is for each master mix to be an equal mix of its constituent libraries.
This is difficult in practice due to the low concentrations and the transfer of small volumes.
In the ideal situation, the concentration of each library would be equal, leading to each library generating an equal number of clusters, and thus an equal number of reads.

As expected, the percentage of clusters from a lane that a library receives is strongly correlated with the number of sequences.

```{r plot-perc}
p_perc_raw <- ggplot(d, aes(x = lane_perc, y = raw_mil)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  labs(x = "Percentage of lane",
       y = "Number of raw reads (millions)",
       title = "Raw")
p_perc_unique <- p_perc_raw %+% aes(y = unique_mil) +
  labs(y = "Number of uniquely mapping reads (millions)",
       title = "Mapped")
plot_grid(p_perc_raw, p_perc_unique)
```

Before combining the libraries into a master mix, they are diluted to the same concentration.
The library quantification is done via a qPCR-based method specifically designed for sequencing libraries.
These estimates are in the variable `qpcr`.
We also measured the diluted libraries, which should all be close to equal.
These are in the variable `qpcr_dilute`.

The concentration of the diluted libraries are not all equal.
Furthermore, they do not correlate well with the percentage of the lane they are sequenced on.
On performing the dilutions, it appears that original libraries with a higher concentrations are systematically overdiluted.

```{r plot-qpcr, warning=FALSE}
p_perc_dilute <- p_perc_raw %+% aes(y = log10(qpcr_dilute)) +
  labs(y = "Conc. of diluted libraries (log10)",
       title = "")
p_perc_qpcr <- p_perc_raw %+% aes(y = log10(qpcr)) +
  labs(y = "Conc. of original libraries (log10)",
       title = "")
p_qpcr_dilute <- p_perc_qpcr %+% aes(x = log10(qpcr_dilute)) +
  labs(x = "Conc. of original libraries (log10)",
       y = "Conc. of diluted libraries (log10)",
       title = "")
plot_grid(p_perc_dilute, p_perc_qpcr, p_qpcr_dilute)
```

But it is not the absolute concentration that matters, but the relative concentrations within each master mix.
This relative concentration should correlate better with the percentage of the lane.

```{r relative-concentration}
d <- d %>%
  group_by(master_mix) %>%
  mutate(qpcr_relative = qpcr_dilute / sum(qpcr_dilute, na.rm = TRUE) * 100)
relative_totals <- tapply(d$qpcr_relative, d$master_mix, sum, na.rm = TRUE)
stopifnot(relative_totals > 99, relative_totals < 101, !is.na(relative_totals))
```

This shows a much better correlation, but still far from perfect.

```{r plot-relative, warning=FALSE}
p_perc_rel <- p_perc_raw %+% d %+% aes(y = qpcr_relative) +
  labs(y = "Relative concentration")
p_perc_rel
```

## Which master mixes need resequenced and possibly rebalanced?

```{r summarize-master-mix}
d_mm <- d %>%
  group_by(master_mix) %>%
  summarize(raw_min = min(raw_mil),
            raw_mean = mean(raw_mil),
            raw_median = median(raw_mil),
            raw_max = max(raw_mil),
            perc_min = min(lane_perc),
            perc_mean = mean(lane_perc),
            perc_median = median(lane_perc),
            perc_max = max(lane_perc))
stopifnot(!is.na(d_mm))
d_mm$master_mix <- reorder(d_mm$master_mix,
                           as.numeric(sub("m", "", d_mm$master_mix)))
```

The average is below 45 million raw reads for each master mix, so they should all be sequenced again.

```{r raw-per-master-mix}
ggplot(d_mm, aes(x = master_mix)) +
  geom_point(aes(y = raw_mean), color = "blue") +
  geom_errorbar(aes(ymax = raw_max, ymin = raw_min)) +
  geom_hline(yintercept = 45, col = "red") +
  labs(x = "Master mix",
       y = "Mean raw reads (millions)\nrange is min to max",
       title = "Raw reads per master mix")
```

What is the distribution of lane percentages for each master mix?

```{r lane-perc-master-mix}
ggplot(d_mm, aes(x = master_mix)) +
  geom_point(aes(y = perc_mean), color = "blue") +
  geom_errorbar(aes(ymax = perc_max, ymin = perc_min)) +
  geom_hline(yintercept = 25, col = "red") +
  geom_hline(yintercept = 5, col = "blue") +
  labs(x = "Master mix",
       y = "Mean lane percentage\nrange is min to max",
       title = "Lane percentage per master mix")
```

On the high end, master mixes 17 and 23 each have a sample that accounted for more than 25% of the sequencing.

```{r too-high}
d_mm %>% filter(perc_max > 25) 
```

The low end is more concerning, with 7 libraries with a minumum lane percentage lower than 1%.

```{r too-low}
d_mm %>% filter(perc_min < 1)
```

## Plan

The core has been sequencing two master mixes to a lane.
How many of these combined pools are ready for sequencing without rebalancing?

```{r filter}
d_mm <- d_mm %>%
  mutate(mix_num = as.numeric(sub("m", "", master_mix))) %>%
  arrange(mix_num) %>%
  mutate(pool = rep(1:12, each = 2)[-nrow(d_mm)])
d_mm %>%
  filter(perc_min > 1 & perc_max < 25) %>%
  select(master_mix, mix_num, pool)
```

The pools 9+10, 13+14, 15+16, and 19+20 are ready for sequencing now.
I'll tell the core to sequence these four lanes, and I'll have Claudia rebalance the remaining 15 master mixes.

Select the master mixes to be rebalanced.

```{r rebalance}
# master mixes to be rebalanced
rebalance <- d_mm %>%
  filter(!(pool %in% c(5, 7, 8, 10))) %>%
  select(master_mix) %>%
  unlist %>% as.character
stopifnot(length(rebalance) == 23 - 8)
```

Calculate the relative lane percentage.
Can't use exact number because two master mixes were sequenced per lane in most cases.
For the few samples with 0% of the lane, change it so that they contributed 0.5%.

```{r relative-lane-perc}
d_out <- d %>%
  mutate(lane_perc = ifelse(lane_perc > 0, lane_perc, 0.5)) %>%
  group_by(master_mix) %>%
  mutate(lane_relative = lane_perc / sum(lane_perc)) %>%
  ungroup
lane_relative_totals <- tapply(d_out$lane_relative, d_out$master_mix, sum)
stopifnot(lane_relative_totals > .99, lane_relative_totals < 1.01,
          !is.na(lane_relative_totals))
```

Need to calculate the volume of each diluted library to add for the new master mixes.
In order to balance the number of molecules contributed by each library, a higher volume of the libraries with a low lane percentage will need to be added compared to those with a higher lane percentage.
For example, if 1 uL of a library which had 30% of the lane was added, then a library which had 10% of the lane would need 3 uL added.
To make it feasible to pipette, I'll make the volume of the library with the highest lane percentage be 1 uL.
Since all the master mixes have different combinations of lane percentages, the final volume of each new master mix will be different.

```{r calc-volume}
calc_volume <- function(fraction, min_volume, digits = 2) {
  # fraction - relative fraction of lane (must sum to 1)
  # min_volume - volume (uL) of sample with largest fraction of lane
  # digits - number of decimal places for rounding volume estimates
  total_perc <- sum(fraction)
  stopifnot(total_perc > .99, total_perc < 1.01)
  max_index <- which.max(fraction)
  volume <- fraction[max_index] / fraction * min_volume
  volume <- round(volume, digits = digits)
  return(volume)
}
# Test function
test_fraction <- c(0.5, 0.25, 0.1, 0.1, 0.05)
stopifnot(calc_volume(test_fraction, min_volume = 1) == c(1, 2, 5, 5, 10))
```

```{r calc-volume-per-mm}
d_out <- d_out %>%
  group_by(master_mix) %>%
  mutate(volume = calc_volume(lane_relative, min_volume = 1)) %>%
  ungroup
```

Format the output to correspond to the spreadsheet containing the experimental data.

```{r format-output}
out <- d_out %>%
  filter(master_mix %in% rebalance) %>%
  mutate(lib_num = as.numeric(sub("s", "", sample)),
         mix_num = as.numeric(sub("m", "", master_mix)),
         lane_relative = round(lane_relative, digits = 4)) %>%
  arrange(lib_num) %>%
  select(mix_num, lib_num, lane_relative, volume)
out %>% filter(mix_num == 1)
write.table(out, file = "../data/rebalancing.txt", quote = FALSE,
            row.names = FALSE, sep = "\t",
            col.names = c("MM", "lib sample #", "fraction", "volume (uL)"))
```

## Session information

```{r info}
sessionInfo()
```
